\chapter{Declaration}

With the exception of chapters \ref{ch:chap_1}, \ref{ch:chap_2} and \ref{ch:chap_3}, which contain introductory material, 
all work in this thesis was primarily carried out by the author under the supervision
of Prof. Ik Siong Heng and Dr. Chris Messenger unless otherwise explicitly stated. 

The work carried out in Ch.~\ref{ch:chap_4}, was primarily 
done by myself under the supervision of Dr. Chris Messenger 
and Prof. Ik Siong Heng, along with support from co-authors 
Mr. Fergus Hayes and Mr. Michael Williams. Both Michael Williams 
and Fergus Hayes had carried out initial studies and produced some 
intial code to work from. Michael Williams, along with myself, 
produced the machine learning code used in the analysis. I, along with input from 
Chris Messenger, produced the matched filtering code used in the 
analysis. All co-authors 
on the published paper version of this analysis~\cite{PhysRevLett.120.141103} 
have also contributed to the text of the published manuscript. 
Both Chris Messenger and Ik Siong Heng provided advice 
on the thesis chapter version of the published manuscript.

The work carried out in Ch.~\ref{ch:chap_5} was primarily 
lead by myself under the supervision of Dr. Chris Messenger 
and Prof. Ik Siong Heng, along with support from co-authors 
Mr. Francesco Tonolini and Prof. Roderick Murray-Smith. Both 
Francesco Tonolini and Roderick Murray-Smith provided guidance 
on the machine learning methods used, some initial machine learning code to work from, 
as well as input 
on the text released in the public version of the manuscript
found here~\cite{1909.06296}. Chris Messenger and Ik Siong Heng also provided 
input to the text in the public form of the manuscript, as well as advice on 
the thesis chapter version.

The final conclusions chapter (Ch.~\ref{ch:chap_6}) was written by myself, with advice 
from Chris Messenger and Ik Siong Heng.
